---
title: "Sleep, Screen Time and Stress Analysis Using Machine Learning Methods"
author: "Nandanie Singh & Inusah Bihinaa Mohammed"
date: "May 8, 2026"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

library(tidyverse)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(class)
library(glmnet)
library(pROC)
library(knitr)
library(reshape2)
library(gridExtra)

set.seed(123)
```

# Load Data
```{r} 
# Load dataset
stress_data <- read_csv("FDAdataset.csv")

# View structure
str(stress_data)
summary(stress_data)

```

# Prepare Data 
```{r}
# Restrict dataset to individuals ages 18 to 45
# and randomly select 10,000 observations
set.seed(123)

stress_data <- stress_data %>%
  filter(age >= 18, age <= 45) %>%
  sample_n(10000)
```

```{r}
# Convert categorical variables to factors
stress_data <- stress_data %>%
  mutate(
    gender = as.factor(gender),
    occupation = as.factor(occupation)
  )

# Remove user_id because it is only an identifier
stress_data <- stress_data %>%
  select(-user_id)

# Create interaction term between sleep duration and screen time
stress_data <- stress_data %>%
  mutate(
    sleep_screen_interaction = sleep_duration_hours * daily_screen_time_hours
  )

# Check missing values
colSums(is.na(stress_data))
```

# Summary Statistics 
```{r}
# Summary statistics for all numeric variables
summary(stress_data %>% select(where(is.numeric)))

# Frequency table for gender
table(stress_data$gender)

# Frequency table for occupation
table(stress_data$occupation)
```

# Correlation Analysis 
```{r}
# Select numeric variables for correlation analysis
numeric_vars <- stress_data %>%
  select(where(is.numeric))

# Create correlation matrix
cor_matrix <- cor(numeric_vars, use = "complete.obs")

# View rounded correlation matrix
round(cor_matrix, 2)

# Convert correlation matrix to long format
cor_long  <- melt(cor_matrix)

# Correlation Heatmap
ggplot(cor_long,
       aes(x = Var1,
           y = Var2,
           fill = value)) +

  geom_tile(color = "white") +

  # Add correlation values
  geom_text(aes(label = round(value, 2)),
            size = 3) +

  scale_fill_gradient2(
    low = "white",
    mid = "orange",
    high = "darkred",
    midpoint = 0,
    limit = c(-1, 1),
    name = "Correlation"
  ) +

  labs(
    title = "Correlation Heatmap of Study Variables",
    x = "",
    y = ""
  ) +

  theme_minimal(base_size = 12) +

  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    panel.grid = element_blank(),
    plot.title = element_text(
      face = "bold",
      hjust = 0.5
    )
  )

# Correlations with Stress Level

# Select numeric variables
numeric_vars <- stress_data %>%
  select(where(is.numeric))

# Calculate correlations with stress level
stress_cor <- cor(numeric_vars)[, "stress_level"]

# Convert to dataframe
stress_cor_table <- data.frame(
  Variable = names(stress_cor),
  Correlation = as.numeric(stress_cor)
)

# Remove stress level correlating with itself
stress_cor_table <- stress_cor_table %>%
  filter(Variable != "stress_level")

# Sort by absolute correlation
stress_cor_table <- stress_cor_table %>%
  mutate(abs_cor = abs(Correlation)) %>%
  arrange(desc(abs_cor))

# Keep strongest relationships
stress_cor_table <- stress_cor_table %>%
  select(Variable, Correlation)

# Display table
kable(
  stress_cor_table,
  digits = 2,
  caption = "Correlations Between Stress Level and Study Variables"
)
```

# Exploratory Figures 
```{r} 
# Create stress group for EDA only
eda_data <- stress_data %>%
  mutate(
    stress_group = ifelse(
      stress_level > median(stress_level, na.rm = TRUE),
      "High Stress",
      "Low Stress"
    ),
    stress_group = factor(stress_group, levels = c("Low Stress", "High Stress"))
  )

# Plot stress outcome
p1 <- ggplot(eda_data, aes(x = stress_group, fill = stress_group)) +
  geom_bar(color = "black") +
  labs(
    title = "Distribution of Stress Outcome",
    x = "Stress Status",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Plot sleep duration
p2 <- ggplot(eda_data, aes(x = sleep_duration_hours)) +
  geom_histogram(bins = 30, color = "black") +
  labs(
    title = "Distribution of Sleep Duration",
    x = "Sleep Duration",
    y = "Frequency"
  ) +
  theme_minimal()

# Plot screen time
p3 <- ggplot(eda_data, aes(x = daily_screen_time_hours)) +
  geom_histogram(bins = 30, color = "black") +
  labs(
    title = "Distribution of Screen Time",
    x = "Daily Screen Time",
    y = "Frequency"
  ) +
  theme_minimal()

# Plot mental fatigue
p4 <- ggplot(eda_data, aes(x = mental_fatigue_score)) +
  geom_histogram(bins = 30, color = "black") +
  labs(
    title = "Distribution of Mental Fatigue",
    x = "Mental Fatigue Score",
    y = "Frequency"
  ) +
  theme_minimal()

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

# Train-Test Split 
```{r} 
# Split data into training and testing sets
set.seed(123)

train_index <- sample(
  seq_len(nrow(stress_data)),
  size = 0.8 * nrow(stress_data)
)

train_data <- stress_data[train_index, ]
test_data <- stress_data[-train_index, ]
```

# Create Binary Stress Outcome 
```{r}
# Use training-set median as cutoff
cutoff <- median(train_data$stress_level, na.rm = TRUE)

# Create binary stress outcome in training data
train_data <- train_data %>%
  mutate(
    stress_binary = ifelse(
      stress_level > cutoff,
      "High Stress",
      "Low Stress"
    ),
    stress_binary = factor(
      stress_binary,
      levels = c("Low Stress", "High Stress")
    )
  )

# Create binary stress outcome in testing data
test_data <- test_data %>%
  mutate(
    stress_binary = ifelse(
      stress_level > cutoff,
      "High Stress",
      "Low Stress"
    ),
    stress_binary = factor(
      stress_binary,
      levels = c("Low Stress", "High Stress")
    )
  )

# Check stress group balance
table(train_data$stress_binary)
table(test_data$stress_binary)
```

# Logistic Regression 
```{r} 
# Fit logistic regression model
log_model <- glm(
  stress_binary ~ sleep_duration_hours +
    daily_screen_time_hours +
    age +
    caffeine_intake_cups +
    sleep_quality_score +
    phone_usage_before_sleep_minutes +
    physical_activity_minutes +
    notifications_received_per_day +
    mental_fatigue_score +
    gender +
    occupation +
    sleep_screen_interaction,
  family = binomial,
  data = train_data
)

summary(log_model)

# Predict probabilities
log_prob <- predict(
  log_model,
  newdata = test_data,
  type = "response"
)

# Convert probabilities into predicted classes
log_pred <- ifelse(
  log_prob > 0.5,
  "High Stress",
  "Low Stress"
)

log_pred <- factor(
  log_pred,
  levels = c("Low Stress", "High Stress")
)

# Evaluate logistic regression
log_cm <- confusionMatrix(
  log_pred,
  test_data$stress_binary,
  positive = "High Stress"
)

log_cm

# AUC
roc_log <- roc(
  test_data$stress_binary,
  log_prob,
  levels = c("Low Stress", "High Stress")
)

auc(roc_log)

# Brier score
actual_numeric <- ifelse(test_data$stress_binary == "High Stress", 1, 0)
log_brier <- mean((log_prob - actual_numeric)^2)
log_brier

# Logistic Regression Coefficients Table
log_results <- data.frame(
  Variable = rownames(summary(log_model)$coefficients),
  Estimate = round(summary(log_model)$coefficients[,1], 3),
  Std_Error = round(summary(log_model)$coefficients[,2], 3),
  P_Value = round(summary(log_model)$coefficients[,4], 4)
)

kable(
  log_results,
  caption = "Logistic Regression Results"
)
```

# Decision Tree 
```{r} 
# Fit decision tree model
tree_model <- train(
  stress_binary ~ sleep_duration_hours +
    daily_screen_time_hours +
    age +
    caffeine_intake_cups +
    sleep_quality_score +
    phone_usage_before_sleep_minutes +
    physical_activity_minutes +
    notifications_received_per_day +
    mental_fatigue_score +
    gender +
    occupation +
    sleep_screen_interaction,
  data = train_data,
  method = "rpart",
  trControl = trainControl(method = "cv", number = 10)
)

tree_model

# Predict classes
tree_pred <- predict(
  tree_model,
  newdata = test_data
)

# Predict probabilities
tree_prob <- predict(
  tree_model,
  newdata = test_data,
  type = "prob"
)[, "High Stress"]

# Evaluate decision tree
tree_cm <- confusionMatrix(
  tree_pred,
  test_data$stress_binary,
  positive = "High Stress"
)

tree_cm

# AUC
roc_tree <- roc(
  test_data$stress_binary,
  tree_prob,
  levels = c("Low Stress", "High Stress")
)

auc(roc_tree)

# Brier score
tree_brier <- mean((tree_prob - actual_numeric)^2)
tree_brier
```

# Random Forest 
```{r} 
# Fit random forest model
set.seed(123)

rf_model <- randomForest(
  stress_binary ~ sleep_duration_hours +
    daily_screen_time_hours +
    age +
    caffeine_intake_cups +
    sleep_quality_score +
    phone_usage_before_sleep_minutes +
    physical_activity_minutes +
    notifications_received_per_day +
    mental_fatigue_score +
    gender +
    occupation +
    sleep_screen_interaction,
  data = train_data,
  ntree = 100,
  importance = TRUE
)

# Predict classes
rf_pred <- predict(
  rf_model,
  newdata = test_data
)

# Predict probabilities
rf_prob <- predict(
  rf_model,
  newdata = test_data,
  type = "prob"
)[, "High Stress"]

# Evaluate random forest
rf_cm <- confusionMatrix(
  rf_pred,
  test_data$stress_binary,
  positive = "High Stress"
)

rf_cm

# AUC
roc_rf <- roc(
  test_data$stress_binary,
  rf_prob,
  levels = c("Low Stress", "High Stress")
)

auc(roc_rf)

# Brier score
rf_brier <- mean((rf_prob - actual_numeric)^2)
rf_brier

# Variable importance
importance(rf_model)
varImpPlot(rf_model, main = "Random Forest Variable Importance")
```

# KNN 
```{r} 
# Prepare data for KNN
train_knn <- train_data %>%
  select(-stress_level, -stress_binary)

test_knn <- test_data %>%
  select(-stress_level, -stress_binary)

# Create dummy variables for categorical predictors
dummy_model <- dummyVars(~ ., data = train_knn)

train_x <- predict(dummy_model, newdata = train_knn)
test_x <- predict(dummy_model, newdata = test_knn)

# Scale data using training data values
train_x <- scale(train_x)

test_x <- scale(
  test_x,
  center = attr(train_x, "scaled:center"),
  scale = attr(train_x, "scaled:scale")
)

# Fit KNN model
knn_pred <- knn(
  train = train_x,
  test = test_x,
  cl = train_data$stress_binary,
  k = 5
)

knn_pred <- factor(
  knn_pred,
  levels = c("Low Stress", "High Stress")
)

# Evaluate KNN
knn_cm <- confusionMatrix(
  knn_pred,
  test_data$stress_binary,
  positive = "High Stress"
)

knn_cm
```

# Lasso Logistic Regression 
```{r} 
# Create model matrix for LASSO
x_train <- model.matrix(
  stress_binary ~ sleep_duration_hours +
    daily_screen_time_hours +
    age +
    caffeine_intake_cups +
    sleep_quality_score +
    phone_usage_before_sleep_minutes +
    physical_activity_minutes +
    notifications_received_per_day +
    mental_fatigue_score +
    gender +
    occupation +
    sleep_screen_interaction,
  data = train_data
)[, -1]

x_test <- model.matrix(
  stress_binary ~ sleep_duration_hours +
    daily_screen_time_hours +
    age +
    caffeine_intake_cups +
    sleep_quality_score +
    phone_usage_before_sleep_minutes +
    physical_activity_minutes +
    notifications_received_per_day +
    mental_fatigue_score +
    gender +
    occupation +
    sleep_screen_interaction,
  data = test_data
)[, -1]

# Convert outcome to numeric for glmnet
y_train <- ifelse(train_data$stress_binary == "High Stress", 1, 0)

# Fit LASSO model
set.seed(123)

lasso_model <- cv.glmnet(
  x = x_train,
  y = y_train,
  family = "binomial",
  alpha = 1
)

# Best lambda
lasso_model$lambda.min

# Predict probabilities
lasso_prob <- predict(
  lasso_model,
  s = "lambda.min",
  newx = x_test,
  type = "response"
)

lasso_prob <- as.numeric(lasso_prob)

# Convert probabilities into predicted classes
lasso_pred <- ifelse(
  lasso_prob > 0.5,
  "High Stress",
  "Low Stress"
)

lasso_pred <- factor(
  lasso_pred,
  levels = c("Low Stress", "High Stress")
)

# Evaluate LASSO
lasso_cm <- confusionMatrix(
  lasso_pred,
  test_data$stress_binary,
  positive = "High Stress"
)

lasso_cm

# AUC
roc_lasso <- roc(
  test_data$stress_binary,
  lasso_prob,
  levels = c("Low Stress", "High Stress")
)

auc(roc_lasso)

# Brier score
lasso_brier <- mean((lasso_prob - actual_numeric)^2)
lasso_brier

# View LASSO coefficients
coef(lasso_model, s = "lambda.min")
```

# Ridge Regression
```{r}
# Ridge uses alpha = 0
set.seed(123)

ridge_model <- cv.glmnet(
  x = x_train,
  y = y_train,
  family = "binomial",
  alpha = 0
)

# Predict probabilities
ridge_prob <- predict(
  ridge_model,
  s = "lambda.min",
  newx = x_test,
  type = "response"
)

ridge_prob <- as.numeric(ridge_prob)

# Convert probabilities into predicted classes
ridge_pred <- ifelse(
  ridge_prob > 0.5,
  "High Stress",
  "Low Stress"
)

ridge_pred <- factor(
  ridge_pred,
  levels = c("Low Stress", "High Stress")
)

# Evaluate Ridge
ridge_cm <- confusionMatrix(
  ridge_pred,
  test_data$stress_binary,
  positive = "High Stress"
)

ridge_cm

# AUC
roc_ridge <- roc(
  test_data$stress_binary,
  ridge_prob,
  levels = c("Low Stress", "High Stress")
)

auc(roc_ridge)

# Brier score
ridge_brier <- mean((ridge_prob - actual_numeric)^2)
ridge_brier
```

# Elastic Net Logistic Regression
```{r}

# Elastic Net uses alpha between 0 and 1
# Here, alpha = 0.5 gives a balance between Ridge and LASSO
set.seed(123)

elastic_model <- cv.glmnet(
  x = x_train,
  y = y_train,
  family = "binomial",
  alpha = 0.5
)

# Predict probabilities
elastic_prob <- predict(
  elastic_model,
  s = "lambda.min",
  newx = x_test,
  type = "response"
)

elastic_prob <- as.numeric(elastic_prob)

# Convert probabilities into predicted classes
elastic_pred <- ifelse(
  elastic_prob > 0.5,
  "High Stress",
  "Low Stress"
)

elastic_pred <- factor(
  elastic_pred,
  levels = c("Low Stress", "High Stress")
)

# Evaluate Elastic Net
elastic_cm <- confusionMatrix(
  elastic_pred,
  test_data$stress_binary,
  positive = "High Stress"
)

elastic_cm

# AUC
roc_elastic <- roc(
  test_data$stress_binary,
  elastic_prob,
  levels = c("Low Stress", "High Stress")
)

auc(roc_elastic)

# Brier score
elastic_brier <- mean((elastic_prob - actual_numeric)^2)
elastic_brier

```

# Model Comparison 
```{r} 
# Create final model comparison table
model_results <- data.frame(
  Model = c(
    "Logistic Regression",
    "Decision Tree",
    "Random Forest",
    "KNN",
    "LASSO",
    "Ridge",
    "Elastic Net"
  ),
  Accuracy = c(
    log_cm$overall["Accuracy"],
    tree_cm$overall["Accuracy"],
    rf_cm$overall["Accuracy"],
    knn_cm$overall["Accuracy"],
    lasso_cm$overall["Accuracy"],
    ridge_cm$overall["Accuracy"],
    elastic_cm$overall["Accuracy"]
  ),
  Sensitivity = c(
    log_cm$byClass["Sensitivity"],
    tree_cm$byClass["Sensitivity"],
    rf_cm$byClass["Sensitivity"],
    knn_cm$byClass["Sensitivity"],
    lasso_cm$byClass["Sensitivity"],
    ridge_cm$byClass["Sensitivity"],
    elastic_cm$byClass["Sensitivity"]
  ),
  Specificity = c(
    log_cm$byClass["Specificity"],
    tree_cm$byClass["Specificity"],
    rf_cm$byClass["Specificity"],
    knn_cm$byClass["Specificity"],
    lasso_cm$byClass["Specificity"],
    ridge_cm$byClass["Specificity"],
    elastic_cm$byClass["Specificity"]
  ),
  F1 = c(
    log_cm$byClass["F1"],
    tree_cm$byClass["F1"],
    rf_cm$byClass["F1"],
    knn_cm$byClass["F1"],
    lasso_cm$byClass["F1"],
    ridge_cm$byClass["F1"],
    elastic_cm$byClass["F1"]
  ),
  AUC = c(
    auc(roc_log),
    auc(roc_tree),
    auc(roc_rf),
    NA,
    auc(roc_lasso),
    auc(roc_ridge),
    auc(roc_elastic)
  ),
  Brier = c(
    log_brier,
    tree_brier,
    rf_brier,
    NA,
    lasso_brier,
    ridge_brier,
    elastic_brier
  )
)

# Display table
kable(
  model_results,
  digits = 2,
  caption = "Model Performance Comparison",
  col.names = c(
    "Model",
    "Acc",
    "Sens",
    "Spec",
    "F1",
    "AUC",
    "Brier"
  )
)

kable(
  model_results,
  digits = 3,
  caption = "Model Performance Comparison"
)

```

```{r}
# Create plotting dataset
plot_results <- model_results %>%
  select(Model, Accuracy, AUC) %>%
  pivot_longer(
    cols = c(Accuracy, AUC),
    names_to = "Metric",
    values_to = "Value"
  )

# Plot model performance
ggplot(plot_results,
       aes(x = Model,
           y = Value,
           fill = Metric)) +
  geom_bar(
    stat = "identity",
    position = "dodge",
    color = "black"
  ) +
  labs(
    title = "Model Performance Comparison",
    x = "Model",
    y = "Performance Value"
  ) +
  ylim(0, 1) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1)
  )
```

# ROC Curves 
```{r} 
# Plot logistic regression ROC
plot(
  roc_log,
  col = "blue",
  lwd = 2,
  main = "ROC Curves for Classification Models"
)

# Add other ROC curves
lines(roc_tree, col = "red", lwd = 2)
lines(roc_rf, col = "darkgreen", lwd = 2)
lines(roc_lasso, col = "purple", lwd = 2)
lines(roc_ridge, col = "orange", lwd = 2)
lines(roc_elastic, col = "brown", lwd = 2)

# Add diagonal reference line
abline(a = 0, b = 1, lty = 2, col = "gray")

# Add legend
legend(
  "bottomright",
  legend = c(
    "Logistic Regression",
    "Decision Tree",
    "Random Forest",
    "LASSO",
    "Ridge",
    "Elastic Net"
  ),
  col = c(
    "blue",
    "red",
    "darkgreen",
    "purple",
    "orange",
    "brown"
  ),
  lwd = 2,
  cex = 0.8
)
```

# Screen Time and Stress Relationship
```{r fig.width=8, fig.height=6}

ggplot(stress_data,
       aes(x = daily_screen_time_hours,
           y = stress_level)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Relationship Between Screen Time and Stress",
    x = "Daily Screen Time (Hours)",
    y = "Stress Level"
  ) +
  theme_minimal()
```


# Sleep Duration and Stress
```{r fig.width=8, fig.height=6}

ggplot(stress_data,
       aes(x = sleep_duration_hours,
           y = stress_level)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Figure 6. Relationship Between Sleep Duration and Stress",
    x = "Sleep Duration (Hours)",
    y = "Stress Level"
  ) +
  theme_minimal()
```

# Predicted Probability Plot 
```{r}
test_data$rf_prob <- rf_prob

ggplot(test_data, aes(x = rf_prob, fill = stress_binary)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  labs(
    title = "Predicted Stress Probability from Random Forest",
    x = "Predicted Probability of High Stress",
    y = "Count",
    fill = "Actual Stress Status"
  ) +
  theme_minimal()
```


Save Results 
```{r} 
write.csv(model_results, "model_results.csv", row.names = FALSE)
```



